Near Optimal Behavior via Approximate State Abstraction

### 2016-05-23 paper/mdp_background.tex
adding changes for camera ready: ferns added, lemma 3 and 4 beefed up, vareps added to functions

The objective of an agent is to solve for the policy that maximizes its expected discounted reward from any state, denoted $\pi^*$. We denote the expected discounted reward for following policy $\pi$ from state $s$ as the value of the state under that policy, $V^\pi(s)$. We similarly denote the expected discounted reward for taking action $a \in \mathcal{A}$ and then following policy $\pi$ from state $s$ forever after as $Q^\pi(s,a)$, defined by the Bellman Equation as:
\begin{equation}
Q^\pi(s,a) = \mathcal{R}(s,a) + \gamma \sum_{s'} \mathcal{T}(s,a,s') Q^\pi(s',\pi(s)).
\end{equation}
Similarly, the value function defined under a given policy, denoted $V^\pi(s)$, is defined as:
\begin{equation}
V^\pi(s) = Q^\pi(s,\pi(s)).
\end{equation}
Lastly, we denote the value and $Q$ functions under the optimal policy as $V^*$ or $V^{\pi^*}$ and $Q^*$ or $Q^{\pi^*}$, respectively. For further background, see~\citet{kaelbling1996reinforcement}.

>>>>

The objective of an agent is to solve for the policy that maximizes its expected discounted reward from any state, denoted $\pi^*$. We denote the expected discounted reward for following policy $\pi$ from state $s$ as the value of the state under that policy, $V^\pi(s)$. We similarly denote the expected discounted reward for taking action $a \in \mathcal{A}$ and then following policy $\pi$ from state $s$ forever after as $Q^\pi(s,a)$, defined by the Bellman Equation as:
\begin{equation}
Q^\pi(s,a) = \mathcal{R}(s,a) + \gamma \sum_{s'} \mathcal{T}(s,a,s') Q^\pi(s',\pi(s')).
\end{equation}
We let \textsc{RMax} denote the maximum reward (which is 1), and \textsc{QMax} denote the maximum Q value, which is $\frac{\textsc{RMax}}{1-\gamma}$.

The value function defined under a given policy, denoted $V^\pi(s)$, is defined as:
\begin{equation}
V^\pi(s) = Q^\pi(s,\pi(s)).
\end{equation}
Lastly, we denote the value and $Q$ functions under the optimal policy as $V^*$ or $V^{\pi^*}$ and $Q^*$ or $Q^{\pi^*}$, respectively. For further background, see~\citet{kaelbling1996reinforcement}.

### 2016-05-23 paper/theory_results.tex
adding changes for camera ready: ferns added, lemma 3 and 4 beefed up, vareps added to functions

We now define types of abstraction based on functions of state--action pairs.
\bdefn{$\ep_f$}
Given a function $f: \mathcal{S}_G \times \mathcal{A} \rightarrow \mathbb{R}$ and a fixed non-negative $\varepsilon \in \mathbb{R}$, we define $\ep_f$ as a type of approximate state-aggregation function that satisfies the following for any two ground states $s_1$ and $s_2$: 
\begin{equation}
\label{eq:phi_f}
\ep_f(s_1) = \ep_f(s_2) \rightarrow \forall_a \left|f(s_1, a) - f(s_2, a)\right| \leq \varepsilon.
\end{equation}
\edefn

That is, when $\ep_f$ aggregates states, all aggregated states have values of $f$ within $\varepsilon$ of each other for all actions.

@@

% Theorem 1
\begin{thm}
There exist at least four types of approximate state-aggregation functions, $\epQ$, $\epM$, $\epB$ and $\epMu$, for which the optimal policy in the abstract \ac{MDP}, applied to the ground \ac{MDP}, has suboptimality bounded polynomially in $\frac{1}{\varepsilon}$:
\begin{equation}
\forall_{s \in \mathcal{S}_G} V_G^{\pi_G^*}(s) - V_G^{\pi_{GA}}(s) \leq poly \left (\frac{1}{\varepsilon} \right).
\end{equation}
\end{thm}

% Subsection: Optimal Q Function
\subsection{Optimal Q Function: $\ep_{Q^*}$}
\label{sec:Q*}

@@

\bdefn{$\epQ$}
An approximate $Q$ function has the same form as Equation~\ref{eq:phi_f}:
\begin{equation}
\ep_{Q^*}(s_1) = \ep_{Q^*}(s_2) \rightarrow \forall_a \left|Q_G^*(s_1, a) - Q_G^*(s_2, a)\right| \leq \varepsilon.
\end{equation}
\edefn

% Q^* Lemma
\begin{lma}
\label{lma:Q*}
When $\ep_{Q^*}$ is used to create the abstract state space $\mcS_A$:
\begin{equation}
\forall_{s \in \mcS_G} V_G^{\pi_G^*}(s) - V_G^{\pi_{GA}}(s) \leq \frac{2\varepsilon}{(1-\gamma)^2}.
\end{equation}
\end{lma}

@@

% Subsection: Model Similarity -- lemma 2
\subsection{Model Similarity: $\ep_{model}$}
\label{sec:model}

@@

%Model Lemma
\begin{lma}
\label{lma:model}
When $\mcS_A$ is created using a function of the $\ep_{model}$ type:
\begin{equation}
\forall_{s \in \mcS_G} V_G^{\pi_G^*}(s) - V_G^{\pi_{GA}}(s) \leq \frac{2\varepsilon + 2\gamma((|\mcS_G|-1)\varepsilon)}{(1-\gamma)^3}.
\end{equation}
\end{lma}

@@

We also assume that the difference in normalizing terms is bounded by some non-negative constant, $k\in \mathbb{R}$, of $\varepsilon$:
\begin{equation}
\left| \sum_b e^{Q_G(s_1,b)} - \sum_b e^{Q_G(s_2,b)} \right| \leq k\times\varepsilon.
\label{eq:bolt_denom}
\end{equation}
\begin{lma} When $S_A$ is created using a function of the $\epB$ type, for some non-negative constant $k \in \mathbb{R}$:
\begin{equation}
\forall_{s \in \mcS_G} V_G^{\pi^*_G}(s) - V_G^{\pi_{GA}}(s) \leq \frac{2\varepsilon\left(\frac{|\mathcal{A}|}{1-\gamma} + k\varepsilon + k\right)}{(1-\gamma)^2}.
\end{equation}
\label{lma:bolt_lemma}
\end{lma}
We use the approximation for $e^x$, with $\delta$ error:
\begin{equation}
 e^x = 1 + x + \delta  \approx 1 + x.
\label{eq:e_to_x_approx}
\end{equation}

@@

By the approximation in Equation~\ref{eq:e_to_x_approx} and the assumption in Equation~\ref{eq:bolt_denom}:
\begin{align*}
\left|\frac{1 + Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}} - \frac{1 + Q_G(s_2,a) + \delta_2}{\sum_j e^{Q_G(s_1,a_j)} \pm k\varepsilon}\right| \leq \varepsilon
\end{align*}
It follows that:
\begin{equation}
\left|Q_G(s_1,a) - Q_G(s_2,a)\right| \leq \varepsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\varepsilon + k \right).
\label{eq:bolt_qs}
\end{equation}
Consequently, we can consider $\epB$ as a variation of $\epQ$, where the difference in $Q$-values is bounded by $\varepsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\varepsilon + k \right)$. Lemma~\ref{lma:bolt_lemma} then follows directly from Lemma~\ref{lma:Q*}.
\qed

>>>>

We now define types of abstraction based on functions of state--action pairs.
\bdefn{$\ep_{f,\varepsilon}$}
Given a function $f: \mathcal{S}_G \times \mathcal{A} \rightarrow \mathbb{R}$ and a fixed non-negative $\varepsilon \in \mathbb{R}$, we define $\ep_{f,\varepsilon}$ as a type of approximate state-aggregation function that satisfies the following for any two ground states $s_1$ and $s_2$: 
\begin{equation}
\label{eq:phi_f}
\ep_{f,\varepsilon}(s_1) = \ep_{f,\varepsilon}(s_2) \rightarrow \forall_a \left|f(s_1, a) - f(s_2, a)\right| \leq \varepsilon.
\end{equation}
\edefn

That is, when $\ep_{f,\varepsilon}$ aggregates states, all aggregated states have values of $f$ within $\varepsilon$ of each other for all actions.

@@

% Theorem 1
\begin{thm}
There exist at least four types of approximate state-aggregation functions, $\epQ$, $\epM$, $\epB$ and $\epMu$, for which the optimal policy in the abstract \ac{MDP}, applied to the ground \ac{MDP}, has suboptimality bounded polynomially in ${\varepsilon}$:
\begin{equation}
\forall_{s \in \mathcal{S}_G} V_G^{\pi_G^*}(s) - V_G^{\pi_{GA}}(s) \leq poly \left ({\varepsilon} \right).
\end{equation}
\end{thm}

We prove this theorem in the following sections by proving bounds on the error of each individual approximate state-aggregation function type.

% Subsection: Optimal Q Function
\subsection{Optimal Q Function: $\ep_{Q^*,\varepsilon}$}
\label{sec:Q*}

@@

\bdefn{$\epQ$}
An approximate $Q$ function has the same form as Equation~\ref{eq:phi_f}:
\begin{equation}
\ep_{Q^*,\varepsilon}(s_1) = \ep_{Q^*,\varepsilon}(s_2) \rightarrow \forall_a \left|Q_G^*(s_1, a) - Q_G^*(s_2, a)\right| \leq \varepsilon.
\end{equation}
\edefn

% Q^* Lemma
\begin{lma}
\label{lma:Q*}
When $\ep_{Q^*,\varepsilon}$ is used to create the abstract state space $\mcS_A$:
\begin{equation}
\forall_{s \in \mcS_G} V_G^{\pi_G^*}(s) - V_G^{\pi_{GA}}(s) \leq \frac{2\varepsilon}{(1-\gamma)^2}.
\end{equation}
\end{lma}

@@

% Subsection: Model Similarity -- lemma 2
\subsection{Model Similarity: $\ep_{model,\varepsilon}$}
\label{sec:model}

@@

%Model Lemma
\begin{lma}
\label{lma:model}
When $\mcS_A$ is created using a function of the $\ep_{model,\varepsilon}$ type:
\begin{equation}
\forall_{s \in \mcS_G} V_G^{\pi_G^*}(s) - V_G^{\pi_{GA}}(s) \leq \frac{2\varepsilon + 2\gamma((|\mcS_G|-1)\varepsilon)}{(1-\gamma)^3}.
\end{equation}
\end{lma}

@@

We also assume that the difference in normalizing terms is bounded by some non-negative constant, $k\in \mathbb{R}$, of $\varepsilon$:
\begin{equation}
\left| \sum_b e^{Q_G(s_1,b)} - \sum_b e^{Q_G(s_2,b)} \right| \leq k\times\varepsilon.
\label{eq:bolt_denom}
\end{equation}
\begin{lma} When $S_A$ is created using a function of the $\epB$ type, for some non-negative constant $k \in \mathbb{R}$:
\begin{equation}
\forall_{s \in \mcS_G} V_G^{\pi^*_G}(s) - V_G^{\pi_{GA}}(s) \leq \frac{2\varepsilon\left(\frac{|\mathcal{A}|}{1-\gamma} + k\varepsilon + k\right)}{(1-\gamma)^2}.
\end{equation}
\label{lma:bolt_lemma}
\end{lma}
We use the approximation for $e^x$, with $\delta$ error:
\begin{equation}
 e^x = 1 + x + \delta  \approx 1 + x.
\label{eq:e_to_x_approx}
\end{equation}
We let $\delta_1$ denote the error in approximating $e^{Q_G(s_1,a)}$, and $\delta_2$ denote the error in approximating $e^{Q_G(s_2,a)}$.

@@

By the approximation in Equation~\ref{eq:e_to_x_approx} and the assumption in Equation~\ref{eq:bolt_denom}:
\begin{align*}
\left|\frac{1 + Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}} - \frac{1 + Q_G(s_2,a) + \delta_2}{\sum_j e^{Q_G(s_1,a_j)} \pm k\varepsilon}\right| \leq \varepsilon
\label{eq:bolt_with_approx}
\end{align*}
There are two cases of interest - either the $1 + x + \delta$ approximation is maximal or minimal in Equation~\ref{eq:bolt_with_approx}. That is, we either have $\sum_j e^{Q_G(s_1,a_j)} + k\varepsilon$, or $\sum_j e^{Q_G(s_1,a_j)} - k\varepsilon$.

First suppose the former. It follows by algebra that:
\begin{equation}
-\varepsilon \leq \frac{1 + Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}} - \frac{1 + Q_G(s_2,a) + \delta_2}{\sum_j e^{Q_G(s_1,a_j)} + k\varepsilon} \leq \varepsilon
\end{equation}

By algebra, we see that:
\begin{multline*}
-\varepsilon \left(k\varepsilon + \sum_j e^{Q_G(s_1,a_j)}\right) - \delta_1 + \delta_2 \leq \\
k\varepsilon\left(\frac{1+Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}}\right) + Q_G(s_1,a) - Q_G(s_2,a) \leq \\
\varepsilon \left(k\varepsilon + \sum_j e^{Q_G(s_1,a_j)}\right) - \delta_1 + \delta_2
\end{multline*}

In the minimal case, it follows that:
\begin{equation}
-\varepsilon \leq \frac{1 + Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}} - \frac{1 + Q_G(s_2,a) + \delta_2}{\sum_j e^{Q_G(s_1,a_j)} - k\varepsilon} \leq \varepsilon
\end{equation}

By similar algebra to the maximal case:
\begin{multline*}
-\varepsilon \left(-k\varepsilon + \sum_j e^{Q_G(s_1,a_j)}\right) - \delta_1 + \delta_2 \leq \\
-k\varepsilon\left(\frac{1+Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}}\right) + Q_G(s_1,a) - Q_G(s_2,a) \leq \\
\varepsilon \left(k\varepsilon + \sum_j e^{Q_G(s_1,a_j)}\right) - \delta_1 + \delta_2
\end{multline*}

From both cases, it follows that:
\begin{equation}
\left|Q_G(s_1,a) - Q_G(s_2,a)\right| \leq \varepsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\varepsilon + k \right).
\label{eq:bolt_qs}
\end{equation}
Consequently, we can consider $\epB$ as a variation of $\epQ$, where the difference in $Q$-values is bounded by $\varepsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\varepsilon + k \right)$. Lemma~\ref{lma:bolt_lemma} then follows directly from Lemma~\ref{lma:Q*}.
\qed
